The Way Forward in Parallelizing Dynamic Languages

### 2014-05-01 talk/icooolps2014/position-paper.tex
fine-grained locking draft

The first obvious candidate to replace the GIL is to use multiple
locks instead of a single global lock. By refining the granularity
of the locking approach, we gain the ability to run code that does
not access the same objects in parallel.

- support of atomic blocks?\\
- hard to get right (deadlocks, performance, lock-granularity)\\
- very hard to get right for a large language\\
- hard to retro-fit, as all existing code assumes GIL semantics\\
- (there are some semantic differences, right? not given perfect lock-placement, but well)
( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )

>>>>

The first obvious candidate to replace the GIL is to use multiple
locks instead of a single global lock. By refining the granularity of
the locking approach, we gain the ability to run code that does not
access the same objects in parallel. What we lose instead is the
simplicity of the GIL approach. With every additional lock, the
likeliness of deadlocks grows, as well as the overhead that acquiring
and releasing locks produces.

Jython\footnote{www.jython.org} is one project that implements an
interpreter for Python on the JVM\footnote{Java Virtual Machine} and
that uses fine-grained locking to correctly synchronize the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks. Since there is no central location, the
complexity of the implementation is quite a bit greater compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too.  One could simply use one lock per library
to avoid this issue.

In the end, fine-grained locking can transparently replace the GIL
and therefore parallelize existing applications without any
changes. It does however not provide a better synchronization
mechanism to the application like e.g. atomic blocks.

%% - support of atomic blocks?\\
%% - hard to get right (deadlocks, performance, lock-granularity)\\
%% - very hard to get right for a large language\\
%% - hard to retro-fit, as all existing code assumes GIL semantics\\
%% - (there are some semantic differences, right? not given perfect lock-placement, but well)
%% ( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )
